Describing the scale and composition of calls for service in Detroit

### 2021-09-01 paper.Rmd
writing

A study from @ratcliffe2021policing in _Crime Science_ sought to describe the complexity and diversity of public demand for police services in Philadelphia, United States. The study was motivated by the recognition that police spend a considerable proportion of deployed resource resolving incidents which fall outside the traditional crime-fighting role of the police. Instead, the origins of public demand for the police can often be attributed to a lack of supply and/or accessibility failure in other organisations, such as those providing support for people requiring (mental) health assistance [@wood2021can; @van2017law]. The study's findings provided data-driven evidence to suggest that the police indeed spend considerable amounts of time dealing with incidents that could be resolved by other -- often more appropriate -- agencies. Comparable findings were subsequently reported in the United States across multiple (anonymised) police jurisdictions [@lum2021can]. In advance of recent contributions, this short paper uses comparable analyses to describe the public demand for police services in Detroit, United States, using open data and open code which can be used to reproduce the findings (see xxxxxxxxxxxxxxxxxxxxxxxxxxxx).

@@

Using computer aided dispatch (CAD) data from the Philadelphia Police Department (PPD) the @ratcliffe2021policing study provided a breakdown of calls for police service during 2019. CAD codes, used to describe calls, of which eighty unique descriptions originated from the public and required officer dispatch, were categorised into six broad classifications: community issues, crime, medical/public health, proactive policing, quality of life, and traffic duties. Recognising that frequency counts of calls for service do not necessarily reflect the amount of consumed police resource, raw counts were adjusted by the time committed by officers to deal with each incident type. Findings indicated that around 55% of officer shift activity time was spent on calls relating to crime, with the remainder of time allocated to incidents involving medical/public health (9%), the community (7%), proactive policing (5%), quality of life (14%) and traffic duties (11%). Focusing on medical/public health incidents, findings also indicated that calls concentrated in particular areas of the city and during particular hours of the day.

@@

The data used for this study covers the city of Detroit in Michigan, United States. The City of Detroit authority publish calls for service data through their open data portal.^[Accessible via https://data.detroitmi.gov/]. The raw data includes both citizen-initiated 911 calls to request police services and officer-initiated calls spanning back to September 2016. In alignment with @ratcliffe2021policing, the data used here is subset for the year 2019 and excludes those calls initiated by an officer. For each incident, the response time and time on the scene are reported. In sum, these comprise the total time officers spend on the incident. Incidents which had a total allocated time of zero minutes (or negative values) were excluded.

Each incident has a _calldescription_ variable which describes the nature of the call. Call descriptions involving administrative duties (e.g., "start of shift information"), completely unknown problems, and non-deployment (e.g., "employee call in / time off") were removed. This left 207 unique call description categories. In the interests of parsimony and ease of interpretation these categories were recoded into 101 broader call descriptions. ^[We refer readers to the corresponding GitHub repository for further details on how call descriptions were combined (xxxxxxxxxxxxxxxxxxxxxxxxxx). Readers unfamiliar with R code can review the reference table entitled 'categorisation summary'.] Each call description was then categorised into the six demand type classifications used in @ratcliffe2021policing. 2.8% of incidents were deemed unclassifiable. For each of the six demand classifications an 'other' incident type was generated for those incidents which consumed less than 0.2% of police time during the year. This left a total of 48 categories for the reported breakdowns, capturing ~267,000 individual public calls for service during 2019.

@@

\newpage
```{r fig1, fig.align = "center", fig.cap = "Proportional breakdown of deployed time spent on each incident category. Graphic inspired by Ratcliffe (2012).", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig1_time_port.png")
```
\newpage

@@

\newpage
```{r fig2, fig.align = "center", fig.cap = "Mean incident counts by day and week, for each demand type.", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig2_time_heat.png")
```
\newpage

@@

Each incident also has corresponding latitude and longitude coordinates to describe the incident location. 2% of incidents were excluded from the spatial descriptive analyses due to incomplete or unknown geographic information. Individual points were aggregated to 1000ft$^2$ grid cells laid over the city to generate basic descriptive heatmaps of counts during the year. These are visualised for each demand classification in Figure 3.

\newpage
```{r fig3, fig.align = "center", fig.cap = "Spatial patterning of total incident counts for each demand type.", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig3_maps.png")
```
\newpage

## Discussion

>>>>

A study from @ratcliffe2021policing in _Crime Science_ sought to describe the complexity and diversity of public demand for police services in Philadelphia, United States. The study was motivated by the recognition that police spend a considerable proportion of deployed resource resolving incidents which fall outside the traditional crime-fighting role of the police. Instead, the origins of public demand for the police can often be attributed to a lack of supply and/or accessibility failure in other organisations, such as those providing support for people requiring (mental) health assistance [@wood2021can; @van2017law]. The study's findings provided data-driven evidence to suggest that the police indeed spend considerable amounts of time dealing with incidents that could be resolved by other -- often more appropriate -- agencies. Comparable findings were subsequently reported in the United States across multiple (anonymised) police jurisdictions [@lum2021can]. In advance of recent contributions, this short paper describes the public demand for police services in Detroit, United States, using open data and open code which can be used to reproduce the findings (see xxxxxxxxxxxxxxxxxxxxxxxxxxxx).

@@

Using computer aided dispatch (CAD) data from the Philadelphia Police Department (PPD) the @ratcliffe2021policing study provided a breakdown of calls for police service during 2019. CAD codes, discrete descriptions of the incident based on information available to the dispatcher, of which eighty unique descriptions originated from the public and required officer dispatch, were categorised into six broad classifications: community issues, crime, medical/public health, proactive policing, quality of life, and traffic duties. Recognising that frequency counts of calls for service do not necessarily reflect the amount of consumed police resource, raw counts were adjusted by the time committed by officers to deal with each incident type. Findings indicated that around 55% of officer shift activity time was spent on calls relating to crime, with the remainder of time allocated to incidents involving medical/public health (9%), the community (7%), proactive policing (5%), quality of life (14%) and traffic duties (11%). Focusing on medical/public health incidents, findings also indicated that calls concentrated in particular areas of the city and during particular hours of the day.

@@

The data used for this study covers the city of Detroit in Michigan, United States. The City of Detroit authority publish calls for service data through their open data portal.^[Accessible via https://data.detroitmi.gov/]. The raw data includes both citizen-initiated 911 calls to request police services and officer-initiated calls spanning back to September 2016. In alignment with @ratcliffe2021policing, the data used here is subset for the year 2019 and excludes those calls initiated by an officer. For each incident, the response time and time on the scene are reported. In sum, these comprise the total deployed time officers spend on the incident. Incidents which had a total allocated time of zero minutes (or negative values) were excluded.

Each incident has a _calldescription_ variable which describes the nature of the call. Call descriptions involving administrative duties (e.g., "start of shift information"), completely unknown problems, and non-deployment (e.g., "employee call in / time off") were removed. This left 207 unique call description categories. In the interests of parsimony and ease of interpretation these categories were recoded into 101 broader call descriptions. ^[We refer readers to the corresponding GitHub repository for further details on how call descriptions were combined (xxxxxxxxxxxxxxxxxxxxxxxxxx). Readers unfamiliar with R code can review the reference table entitled 'categorisation summary'.] Each call description was then categorised into the six demand type classifications used in @ratcliffe2021policing. 2.8% of incidents were deemed unclassifiable. For each of the six demand classifications an 'other' incident type was generated for those incidents which consumed less than 0.2% of police time during the year. This left a total of 48 categories for the reported breakdowns, capturing ~267,000 individual public calls for service in Detroit during 2019.

@@

\newpage
```{r fig1, fig.align = "center", fig.cap = "Proportional breakdown of deployed time spent on each incident category. Graphic inspired by Ratcliffe (2021).", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig1_time_port.png")
```
\newpage

@@

The mean incident counts by day of the week and by hour, for each demand classification, are visualised in Figure 2. This demonstrates that the different sources of public demand for police services, as indicated by the six demand classifications, can have distinct temporal patterning. Community calls appear to concentrate on weekdays, around lunchtime. Crime and quality of life-related calls both appear to occur in the evening and early hours of the morning, although crime incidents are more impacted by weekend activity stretching into Saturday and Sunday mornings. Traffic incidents concentrate in the evenings, particularly on a Friday. Proactive policing does not have a distinct hourly patterning, but this could be attributed to the low hourly counts. Calls relating to health issues, the primary interest of @ratcliffe2021policing, appear to largely take place in the evening and early morning, aligning with findings from Philadelphia.

@@

Individual incident location coordinates were aggregated to synthetic grid cells laid over the city to generate heatmaps of counts throughout the year. ^[Grid cells were defined as 1000$^2$, a size deemed appropriate based on a balance between capturing localised variation and interpretable visualisations. 2% of incidents were excluded from the spatial visualisations due to incomplete geographic information.]These are visualised for each demand classification in Figure 3.

\newpage
```{r fig2, fig.align = "center", fig.cap = "Mean incident counts by day and week, for each demand type.", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig2_time_heat.png")
```

```{r fig3, fig.align = "center", fig.cap = "Spatial patterning of total incident counts for each demand type.", out.width = "\\textwidth"}
knitr::include_graphics("visuals/fig3_maps.png")
```
\newpage

# Discussion

# References
